---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
To start, let us set the working directory to the local cloned directory, unzip and load the data.
````{r}
setwd("~/Documents/Data_Science/Reproducible_Research/RepData_PeerAssessment1/")
unzip("activity.zip")
activity<-read.csv("activity.csv")
echo="TRUE"
```
The dates are in "factor" format, we put them to date format with the `as.Date` function.

```{r}
date<-as.Date(as.character(activity$date))
activity$date<-date
```
## What is mean total number of steps taken per day?
In the following chunk of code, we compute the number of total steps per day:
```{r}
totalsteps<-aggregate(activity$steps, list(period=activity$date), sum, na.rm=TRUE)
head(totalsteps)
```
We load the ggplot2 library and make a histogram of the total number of steps per day:
```{r}
library(ggplot2)
```

```{r fig.width=6, fig.height=4}
qplot(as.vector(totalsteps[,2]), xlab = "total number of steps taken each day")
```

We then compute the mean and the median of the number of steps per day:

```{r}
meansteps<-aggregate(activity$steps,list(period=activity$date), mean, na.rm=TRUE)
meansteps<-subset(meansteps,subset=complete.cases(meansteps))
colnames(meansteps)<-c("Days", "Mean Step Number")
row.names(meansteps)<-NULL
mediansteps<-aggregate(activity$steps, list(period=activity$date), median, na.rm=TRUE)
mediansteps<-subset(mediansteps,subset=complete.cases(meansteps))
```
Then, we create the tables for the mean and median values:
```{r kable}
library(knitr)
kable(head(meansteps)); kable(head(mediansteps))
```


## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
